---
title: "Reproducible Research: Peer Assessment 1"
author: Mark Anderson
output: 
  html_document:
    keep_md: true
---

## Introduction ^[https://github.com/rdpeng/RepData_PeerAssessment1]

It is now possible to collect a large amount of data about personal
movement using activity monitoring devices such as a
[Fitbit](http://www.fitbit.com), [Nike
Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or
[Jawbone Up](https://jawbone.com/up). These type of devices are part of
the "quantified self" movement -- a group of enthusiasts who take
measurements about themselves regularly to improve their health, to
find patterns in their behavior, or because they are tech geeks. But
these data remain under-utilized both because the raw data are hard to
obtain and there is a lack of statistical methods and software for
processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring
device. This device collects data at 5 minute intervals through out the
day. The data consists of two months of data from an anonymous
individual collected during the months of October and November, 2012
and include the number of steps taken in 5 minute intervals each day.

## Data ^[https://github.com/rdpeng/RepData_PeerAssessment1]

The data for this assignment can be downloaded from the course web
site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken




The dataset is stored in a comma-separated-value (CSV) file and there
are a total of 17,568 observations in this
dataset.



## Loading and preprocessing the data
### Getting the file
Establish source and destination
```{r echo=TRUE}
sourceurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
file <- "activity.csv"
```
Fetch the file if it doesn't already exist in working directory
```{r echo=TRUE}
if (!file.exists(file)){
  tmpfile <- tempfile()
  download.file(sourceurl,tmpfile,method="curl")
  unzip(tmpfile,file)
  dateDownloaded <- date()
  dateDownloaded
}
```
### import the data
```{r echo=TRUE}
data <- read.csv(file)
str(data)
```
### clean the data
Without casting, data$date is a 'factor'.  Use as.Date to force to 'Date'
```{r echo=TRUE}
data$date <- as.Date(data$date, format="%Y-%m-%d")
str(data)
```
## What is mean total number of steps taken per day?
1. Make a histogram of the total number of steps taken each day
```{r echo=TRUE}
total_steps <- aggregate( steps ~ date, data, sum, na.rm=TRUE)
str(total_steps)
hist(
  total_steps$steps,
  xlab='Steps per day',
  main='Histogram of Steps per day')
```

2. Calculate and report the **mean** and **median** total number of steps taken per day

The mean number of steps per day
```{r echo=TRUE}
mean_steps <- mean(total_steps$steps)
mean_steps
```
The median number of steps per day
```{r echo=TRUE}
median_steps <- median(total_steps$step)
median_steps
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r echo=TRUE}
stepsPerInterval <- aggregate(steps ~ interval, data, FUN = mean, na.rm = TRUE) 
plot(stepsPerInterval, type = "l", col = "red")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r echo=TRUE}
maxStepInterval <- stepsPerInterval[which.max(stepsPerInterval$steps), ]$interval
maxStepInterval
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?

